---
title: "Ecotopic_3_Main_Markdown"
author: "Stellaria Otstot"
date: "2026-03-31"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Research Question

> How does stream oxygen conc vary across time and space?

## Hypotheses

> H1

> H0

> H etc

# Code

## set up

```{r}
library(dataRetrieval)
library(ggplot2)
library(dplyr)
```

## Data Wrangling

### Read in Data

**Site Choice**

> 2 sites from each upper widwest state (MN, MI, WI)
>
> Must have data on:
>
> > Dissolved oxygen, water, unfiltered, milligrams per liter
> >
> > Turbidity, water, unfiltered, monochrome near infra-red LED light,
> > 780-900 nm, detection angle 90 +-2.5 degrees, formazin nephelometric
> > units (FNU)

> *Wisconsin:*
>
> Menominee River Near MC Allister, WI - USGS-04067500
>
> FOX R. CANAL AT LOCK & DAM No. 2 AT KAUKAUNA, WI - USGS-04084485
>
> Milwaukee River at Mouth at Milwaukee, WI - USGS-04087170
>
> Manitowoc River at Manitowoc, WI - USGS-04085427

> *Minnesota*
>
> St. Louis River at Scanlon, MN - USGS-04024000
>
> Red River of the North Near Georgetown, MN - USGS-05062130

> *Michigan*
>
> Mineral River DS of Native CK NR Silver City, MI - USGS-040325155
>
> M BR Escanaba R at US Hwy 41 NR Humboldt, MI - USGS-04057801
>
> AU Sable River Near Glennie, MI - USGS-04137025
>
> Lower River Rouge at Dearborn, MI - USGS-04168400

```{r}
# Find all sites across the us that have water oxygen data 
whatNWISsites()
oxy_sites <- whatNWISsites(stateCd = c("MN", "WI"), parameterCd = "00300") #does not run on stellas R 
```

```{r}
# make site vectors 
Wisconsin_sites <- c("USGS-04067500", "USGS-04084485", "USGS-04087170", "USGS-04085427")
Minnesota_sites <- c("USGS-04024000", "USGS-05062130")
Michigan_sites <- c("USGS-040325155", "USGS-04057801", "USGS-04137025", "USGS-04168400")

# make variable vectors
sites <- c(Wisconsin_sites, Minnesota_sites, Michigan_sites)
param_code <- "00300"
start <- "2010-01-01"
end <- "2025-01-01"
zone = "America/Chicago"

# read in data
waterdat <- read_waterdata_daily(
  monitoring_location_id = sites,
  parameter_code = param_code,
  time = as.POSIXct(c(start, end), 
  tz = zone))
```


### Variable wranging

```{r}
# make seasonal buckets for data
waterdat_cleaned <- waterdat %>%
  mutate(md = format(time, "%m-%d")) %>%
  mutate(season = case_when(md >= "03-01" & md <= "05-31" ~ "Spring",
                            md >= "06-01" & md <= "08-31" ~ "Summer",
                            md >= "09-01" & md <= "11-30" ~ "Fall",
                            TRUE ~"Winter"))
                           

# calculate seasonal means/mins/maxs/range/amplitude

waterdat_cleaned %>%
  group_by(season, monitoring_location_id) %>%
  summarise(mean_o = mean(value)) # mean O level

# make new var that declares whether av/max/min/whatever is above or below healthy threshold (based on lit)
```

# visualizations

```{r}
# plot average water oxygen over time for each site (line/scatter plot)
  ## on this plot have line that corresponds to healthy threshold

waterdat_cleaned %>%
  mutate(location = case_when(monitoring_location_id %in% Michigan_sites ~ "Michigan",
                              monitoring_location_id %in% Minnesota_sites ~ "Minnesota",
                              monitoring_location_id %in% Wisconsin_sites ~ "Wisconsin")) %>%
  group_by(monitoring_location_id, time) %>%
  mutate(mean_o = mean(value)) %>%
  dplyr::distinct(time, .keep_all = TRUE)

# 
```

# future ideas

> make onx map of all sites for presentation

```{r}
library(usmap)
plot_usmap(include = .midwest_region, labels = TRUE) +
  geom_sf(data = waterdat, color = "blue", size = 4, shape = 8)
  

```


### Analysis

```{r}
# water oxygen vs other var
```

# Notes

look at what sites are available for a given parameter characterize the
parameter - for a given site what is the variation of o for a year
(first figure) oxygen at each site over time - is the amplitude across a
year variable across site? - what are healthy ranges of oxygen and how
many sites are above/below the threshold? - what are our hypotheses why
this range? - anova comparing differences between sites (min or av
oxygen (for each season?))

what are potential drivers of this var - correlate two parameters

## start with oxygen and then maybe add in more vars
