---
title: "Ecotopic_3_Main_Markdown"
author: "Stellaria Otstot"
date: "2026-03-31"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Research Question

> How does stream oxygen conc vary across time and space?

## Hypotheses

> H1

> H0

> H etc


# Code

## Data Wrangling

### Read in Data
```{r}
# Find all sites across the us that have water oxygen data
whatNWISsites()
oxy_sites <- whatNWISsites(stateCd = c("MN", "WI"), parameterCd = "00300")


# read in data for these sites (just oxygen var or others too?) 

## day data or continuous data?
```

### Variable wranging
```{r}
# make seasonal buckets for data

# calculate seasonal means/mins/maxs/range/amplitude

# make new var that declares whether av/max/min/whatever is above or below healthy threshold (based on lit)
```

# visualizations
```{r}
# plot average water oxygen over time for each site (line/scatter plot)
  ## on this plot have line that corresponds to healthy threshold

# 
```

# future ideas
```{r}
# water oxygen vs other var
```


# Notes

look at what sites are available for a given parameter
characterize the parameter 
- for a given site what is the variation of o for a year (first figure) oxygen at each site over time
- is the amplitude across a year variable across site? 
- what are healthy ranges of oxygen and how many sites are above/below the threshold?
- what are our hypotheses why this range?
- anova comparing differences between sites (min or av oxygen (for each season?))

what are potential drivers of this var
- correlate two parameters

start with oxygen and then maybe add in more vars 
- 