---
title: "Ecotopic_3_Main_Markdown"
author: "Stellaria Otstot"
date: "2026-03-31"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Research Question

> How does stream oxygen conc vary across time and space?

## Hypotheses

> At a given site, as temperature increases disolved oxygen will decrease because cold water can hold more disolved oxygen than warm water

> Disolved oxygen will be lower during the summer because of increased temperatures during these months

> H etc

# Code

## set up

```{r}
library(dataRetrieval)
library(ggplot2)
library(dplyr)
library(usmap)
library(sf)
library(lubridate)
library(tidyverse)
```

## Data Wrangling

### Read in Data

**Site Choice**

> 2 sites from each upper widwest state (MN, MI, WI)
>
> Must have data on:
>
> > Dissolved oxygen, water, unfiltered, milligrams per liter
> >
> > Turbidity, water, unfiltered, monochrome near infra-red LED light,
> > 780-900 nm, detection angle 90 +-2.5 degrees, formazin nephelometric
> > units (FNU)

> *Wisconsin:*
>
> Menominee River Near MC Allister, WI - USGS-04067500
>
> FOX R. CANAL AT LOCK & DAM No. 2 AT KAUKAUNA, WI - USGS-04084485
>
> Milwaukee River at Mouth at Milwaukee, WI - USGS-04087170
>
> Manitowoc River at Manitowoc, WI - USGS-04085427

> *Minnesota*
>
> St. Louis River at Scanlon, MN - USGS-04024000
>
> Red River of the North Near Georgetown, MN - USGS-05062130

> *Michigan*
>
> Mineral River DS of Native CK NR Silver City, MI - USGS-040325155
>
> M BR Escanaba R at US Hwy 41 NR Humboldt, MI - USGS-04057801
>
> AU Sable River Near Glennie, MI - USGS-04137025
>
> Lower River Rouge at Dearborn, MI - USGS-04168400

# Code

```{r}
# make site vectors 
Wisconsin_sites <- c("USGS-04067500", "USGS-04084485", "USGS-04087170", "USGS-04085427")
Minnesota_sites <- c("USGS-04024000", "USGS-05062130")
Michigan_sites <- c("USGS-040325155", "USGS-04057801", "USGS-04137025", "USGS-04168400")

# make variable vectors
sites <- c(Wisconsin_sites, Minnesota_sites, Michigan_sites)
param_code <- c("00300", "63680", "00010", "00021")
start <- "2010-01-01"
end <- "2025-01-01"
zone = "America/Chicago"

# read in data
waterdat <- read_waterdata_daily(
  monitoring_location_id = sites,
  parameter_code = param_code,
  time = as.POSIXct(c(start, end), 
  tz = zone))

# find all locations with dissolved oxygen
# dO_locations <- unique(unlist(dO_data_avail$monitoring_location_id))
# dO_data <- read_waterdata_daily(monitoring_location_id = dO_locations, 
#                                 parameter_code = "00300")
# unique(unlist(dO_data$monitoring_location_id))

```


### Data Wranging

```{r}
# make seasonal buckets for data
waterdat_cleaned <- waterdat %>%
  mutate(md = format(time, "%m-%d")) %>%
  mutate(season = case_when(md >= "03-01" & md <= "05-31" ~ "Spring",
                            md >= "06-01" & md <= "08-31" ~ "Summer",
                            md >= "09-01" & md <= "11-30" ~ "Fall",
                            TRUE ~"Winter"))
                           

# calculate seasonal means/mins/maxs/range/amplitude
waterdat_cleaned %>%
  filter(statistic_id == "00003") %>%
  group_by(season, monitoring_location_id)
```


```{r}
# name sites and filter for means
states_cleaned <- waterdat_cleaned %>%
  mutate(state = case_when(monitoring_location_id %in% Michigan_sites ~ "Michigan",
                              monitoring_location_id %in% Minnesota_sites ~ "Minnesota",
                              monitoring_location_id %in% Wisconsin_sites ~ "Wisconsin")) %>%
  mutate(site_name = case_when(monitoring_location_id == "USGS-04067500" ~ "Menominee River",
                               monitoring_location_id == "USGS-04084485" ~ "FOX R. CANAL AT LOCK & DAM",
                               monitoring_location_id == "USGS-04087170" ~ "Milwaukee River",
                               monitoring_location_id == "USGS-04085427" ~ "Manitowoc River",
                               monitoring_location_id == "USGS-04024000" ~ "St. Louis River",
                               monitoring_location_id == "USGS-05062130" ~ "Red River",
                               monitoring_location_id == "USGS-040325155" ~ "Mineral River DS",
                               monitoring_location_id == "USGS-04057801" ~ "M BR Escanaba R",
                               monitoring_location_id == "USGS-04137025" ~ "AU Sable River",
                               monitoring_location_id == "USGS-04168400" ~ "Lower River Rouge")) %>%
  group_by(monitoring_location_id, time) %>%
  mutate(year = year(time))
```

```{r}
states_cleaned_wide <- states_cleaned %>%
  select(-unit_of_measure, -last_modified) %>%
  filter(statistic_id == "00003") %>%
  mutate(parameter_name = case_when(parameter_code == "00300" ~ "dissolved_oxygen",
                                    parameter_code == "00010" ~ "temperature",
                                    parameter_code == "63680" ~ "turbidity",
                                    TRUE ~ "other")) %>%
  select(-parameter_code) %>%
  pivot_wider(id_cols = !time_series_id,
              names_from = parameter_name,
              values_from = value,
              values_fn = mean)
```

# Visualizations

## Preliminary, sigle parameter visualizations RQ1
```{r}
p1 <- states_cleaned %>%
  filter(parameter_code == "00300") %>%
  filter(statistic_id == "00003") %>%
  ggplot(aes(x = season, y = value, color = state)) +
  geom_boxplot() +
  geom_hline(yintercept = 5, linetype = "dotted", linewidth = 0.8, color = "red") +
  theme_minimal() +
  ylab("Mean Dissolved Oxygen (mg/l)")

p1

p2 <-states_cleaned %>%
  filter(statistic_id == "00002") %>%
  filter(parameter_code == "00300") %>%
  ggplot(aes(x = season, y = value, color = state)) +
  geom_boxplot() +
  geom_hline(yintercept = 5, linetype = "dotted", linewidth = 0.8, color = "red") +
  theme_minimal() +
  ylab("Min Dissolved Oxygen (mg/l)")
p2

p3 <- states_cleaned %>%
  filter(statistic_id == "00002") %>%
  filter(parameter_code == "00300") %>%
  filter(!(site_name %in% c("M BR Escanaba R", "Mineral River DS", "Red River"))) %>%
  ggplot(aes(x = time, y = value, color = state)) +
  geom_line() +
  geom_hline(yintercept = 5, linetype = "dotted", linewidth = 0.6, color = "red") +
  facet_wrap(~site_name, ncol = 1) +
  theme_minimal() +
  ylab("Mean Dissolved Oxygen (mg/l)")
p3

p4 <- states_cleaned %>%
  filter(statistic_id == "00002") %>%
  filter(parameter_code == "00300") %>%
  filter(site_name %in% c("M BR Escanaba R", "Mineral River DS", "Red River")) %>%
  ggplot(aes(x = time, y = value, color = state)) +
  geom_line() +
  geom_hline(yintercept = 5, linetype = "dotted", linewidth = 0.6, color = "red") +
  facet_wrap(~site_name, ncol = 1) +
  theme_minimal() +
  ylab("Mean Dissolved Oxygen (mg/l)")
p4
```

## multi var visualizations RQ2
```{r}
#### turbidity
# general
p5 <- states_cleaned_wide %>%
  ggplot(aes(y = as.numeric(dissolved_oxygen), x = as.numeric(turbidity))) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  ylab("Disolved Oxygen (mg/l)") +
  xlab("Turbidity (FNU)")
p5

# color by site wrao by state
p6 <- states_cleaned_wide %>%
  filter(state != "Michigan") %>%
  ggplot(aes(y = as.numeric(dissolved_oxygen), x = as.numeric(turbidity), color = site_name)) +
  facet_wrap(~state, ncol = 1) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  ylab("Disolved Oxygen (mg/l)") +
  xlab("Turbidity (FNU)")
p6

#### temp
# general
p7 <- states_cleaned_wide %>%
  ggplot(aes(y = as.numeric(dissolved_oxygen), x = as.numeric(temperature))) +
  geom_point(size = 0.5) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  ylab("Disolved Oxygen (mg/l)") +
  xlab("Temperature (F)")
p7

# color by site wrap by state
p8 <- states_cleaned_wide %>%
  ggplot(aes(y = as.numeric(dissolved_oxygen), x = as.numeric(temperature), color = site_name)) +
  facet_wrap(~state, ncol = 1) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  ylab("Disolved Oxygen (mg/l)") +
  xlab("Temperature (F)")
p8

# no points (less clutter)
p9 <- states_cleaned_wide %>%
  ggplot(aes(y = as.numeric(dissolved_oxygen), x = as.numeric(temperature), color = site_name)) +
  facet_wrap(~state, ncol = 1) +
  geom_smooth(method = "lm", se = FALSE) +
  theme_minimal() +
  ylab("Disolved Oxygen (mg/l)") +
  xlab("Temperature (F)")
p9
```

# Model Building
```{r}
m5 <- lm(dissolved_oxygen ~ turbidity, data = states_cleaned_wide) 
summary(m5)

m6 <- lm(dissolved_oxygen ~ turbidity + site_name , data = states_cleaned_wide) 
summary(m6)

m7 <- lm(dissolved_oxygen ~ temperature , data = states_cleaned_wide) 
summary(m7)
```



## Map of study sites
```{r}
  
water_dat_labeled_sites <- states_cleaned %>%
  filter(statistic_id == "00003") %>%
  mutate(site_name = case_when(monitoring_location_id == "USGS-04067500" ~ "Menominee River",
                               monitoring_location_id == "USGS-04084485" ~ "FOX R. CANAL AT LOCK & DAM",
                               monitoring_location_id == "USGS-04087170" ~ "Milwaukee River",
                               monitoring_location_id == "USGS-04085427" ~ "Manitowoc River",
                               monitoring_location_id == "USGS-04024000" ~ "St. Louis River",
                               monitoring_location_id == "USGS-05062130" ~ "Red River",
                               monitoring_location_id == "USGS-040325155" ~ "Mineral River DS",
                               monitoring_location_id == "USGS-04057801" ~ "M BR Escanaba R",
                               monitoring_location_id == "USGS-04137025" ~ "AU Sable River",
                               monitoring_location_id == "USGS-04168400" ~ "Lower River Rouge"))

map <- plot_usmap(include = .midwest_region, labels = TRUE) +
  geom_sf(data = water_dat_labeled_sites, color = "blue", size = 4, shape = 8) +
  geom_sf_text(data = water_dat_labeled_sites, aes(label = site_name), nudge_y = -50000, color = "red", check_overlap = TRUE)

map + ggtitle("Monitoring Site Locations") + theme(plot.title = element_text(size = 12, face = "bold", hjust = 0.5))
```


# Notes

look at what sites are available for a given parameter characterize the
parameter - for a given site what is the variation of o for a year
(first figure) oxygen at each site over time - is the amplitude across a
year variable across site? - what are healthy ranges of oxygen and how
many sites are above/below the threshold? - what are our hypotheses why
this range? - anova comparing differences between sites (min or av
oxygen (for each season?))

what are potential drivers of this var - correlate two parameters
- start with oxygen and then maybe add in more vars

# Apendix

## Variable importance analysis
```{r}
# read in data
varimp <- read_waterdata_daily(
  monitoring_location_id = sites, 
  time = as.POSIXct(c(start, end), 
  tz = zone))

varimp_clean <- varimp %>%
  filter(statistic_id == "00003") %>%
  st_drop_geometry()
 
varimp_wide <- varimp_clean %>%
  pivot_wider(names_from = parameter_code, values_from = value)
```

# code to find availble data for usgs sites
```{r}
dO_data_avail <- read_waterdata_ts_meta(parameter_code = "00300") 

dO_data_avail <- dO_data_avail %>% 
  filter(computation_identifier == "Mean") %>%
  filter(state_name %in% c("Minnesota", "Wisconsin", "Michigan"))
```

